### 2024-06-11 paper/tex/04-algorithm.tex
corrections

\paragraph{Algorithmic Variables}
To design a mode-correct inference system, we slightly modify the language we operate on.
The entities (terms, types, contexts) that the algorithm manipulates we call \emph{algorithmic}. 
They extend the previously defined declarative terms and types by adding 
\emph{algorithmic type variables} (\aka unification variables). 
The algorithmic variables represent unknown types, 
which cannot be inferred immediately but are promised to be instantiated
as the algorithm proceeds.

@@

\paragraph{Algorithmic Types}
The syntax of algorithmic types extends the declarative syntax by adding
algorithmic variables as new terminals. We add positive algorithmic variables $[[α̂⁺]]$ 
to the syntax of positive types, and negative algorithmic variables $[[α̂⁻]]$ to the 
syntax of negative types. All the constructors of the system can be applied 
to \emph{algorithmic} types, however, algorithmic variables cannot be abstracted by the
quantifiers $[[∀]]$ and $[[∃]]$.

\paragraph{Algorithmic Contexts $[[Ξ]]$ and Well-formedness}
To specify when algorithmic types are well-formed, we define algorithmic
contexts $[[Ξ]]$ as sets of algorithmic variables. Then
$[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ represent the well-formedness judgment of
algorithmic terms defined as expected. Informally, they check that all free
declarative variables are in $[[Γ]]$, and all free algorithmic variables are in
$[[Ξ]]$. Most of the rules are inherited from the well-formedness of
\emph{declarative} types: the declarative variables are checked to belong to the
context $[[Γ]]$, the quantifiers extend the context, type constructors are
well-formed congruently. As we extend the syntax with algorithmic variables, we
also add two base-case rules for them: \ruleref{\ottdruleWFATPUVarLabel} and
\ruleref{\ottdruleWFATNUVarLabel} (see \cref{fig:algo-wf}).

@@

\paragraph{Instantiation Context $[[Θ]]$}
When one instantiates an algorithmic variable, one may only use type variables
\emph{available in its scope}. As such, each algorithmic variable must remember
the context at the moment when it was introduced. In our algorithm, this
information is represented by an \emph{instantiation context} $[[Θ]]$---a set of
pairs associating algorithmic variables and declarative contexts.

\paragraph{Algorithmic Substitution}
We define the algorithmic substitution $[[uσ]]$ as a mapping from algorithmic
variables to \emph{declarative} types. The signature $[[Θ ⊢ uσ : Ξ]]$
specifies the domain and the range of $[[uσ]]$: for each variable $[[α̂±]]$
in $[[Ξ]]$, there exists an corresponding entry in $[[Θ]]$ associating 
$[[α̂±]]$ with a declarative context $[[Γ]]$ such that $[[ [uσ]α̂± ]]$
is well-formed in $[[Γ]]$. In addition, we assume that $[[uσ]]$
acts as the identity on the variables not in $[[Ξ]]$.

\paragraph{Algorithmic Normalization}
Similarly to well-formedness, the normalization of algorithmic types is defined
by extending the declarative definition (\cref{fig:type-nf}) with the
algorithmic variables. To the rules repeating the declarative normalization, we
add rules saying that normalization is trivial on algorithmic variables.

@@

\paragraph{Auxiliary Functions}
\begin{itemize}
  \item
    We define $[[dom(SC)]]$---the domain of constraint set $[[SC]]$---as a set of
    algorithmic variables that it restricts. Similarly, we define $[[dom(Θ)]]$---the
    domain of instantiation context---as a set of algorithmic variables that
    $[[Θ]]$ associates with their contexts.  
  \item We write $[[Θ(α̂±)]]$ to denote the
    declarative context associated with $[[α̂±]]$ in $[[Θ]]$. 
  \item $[[uv(uN)]]$ and $[[uv(uP)]]$ denote the set of free algorithmic variables in $[[uN]]$ and
    $[[uP]]$ respectively.
  \item We write $[[Γ ⊢ Θ]]$ to denote that each
    declarative context associated with an algorithmic variable in $[[Θ]]$ is 
    a subcontext (subset) of $[[Γ]]$.
\end{itemize}

\paragraph{Equivalent Substitutions}
In the proofs of the algorithm correctness, 
we often state or require that two substitutions are equivalent
on a given set of algorithmic variables.
We denote it as $[[Θ ⊢ uσ' ≈ uσ : Ξ]]$ meaning that
for each $[[α̂±]]$ in $[[Ξ]]$, substitutions $[[uσ]]$ and $[[uσ']]$
map $[[α̂±]]$ to types equivalent in the corresponding context:
$[[ Θ(α̂±) ⊢ [uσ']α̂± ≈ [uσ]α̂± ]]$.

\paragraph{Constraint well-formedness and satisfaction}
Suppose that $[[Θ]]$ is an instantiation context. 
We say that constraint set $[[SC]]$ is well-formed in $[[Θ]]$ 
(denoted as $[[Θ ⊢ SC]]$) 
if for every every constrain entry $[[scE ∊ SC]]$
associating a variable $[[α̂±]]$ with a type $[[iP]]$,
this type is well-formed in the corresponding declarative context $[[Θ(α̂±)]]$.

@@

\paragraph{Variables} Rules \ruleref{\ottdruleANVarLabel} and \ruleref{\ottdruleAPVarLabel}
say that if both of the input types are equal declarative variables,
they are subtypes of each other, with no constraints (as there are no algorithmic variables).

\paragraph{Shifts} Rules \ruleref{\ottdruleAShiftDLabel} and
\ruleref{\ottdruleAShiftULabel} cover the downshift and the upshift cases,
respectively. If the input types are constructed by shifts, then the subtyping
can only hold if they are equivalent. This way, the algorithm must find the
instantiations of the algorithmic variables on the left-hand side
such that these instantiations make the left-hand side and the ride-hand side
equivalent. For this purpose, the algorithm invokes the
unification procedure (\cref{sec:unification}) preceded by the normalization of the input types.
It returns the resulting constraints given by the unification algorithm.

\paragraph{Quantifiers}  
Rules \ruleref{\ottdruleAForallLabel} and 
\ruleref{\ottdruleAExistsLabel} are symmetric. 
Declaratively, the quantified variables on the left-hand side must 
be instantiated with types, which are not known beforehand.
We address this problem by algorithmization 
of the quantified variables (see \cref{sec:algo-syntax}).
The rule introduces fresh algorithmic variables
$[[puas]]$ or $[[nuas]]$,
puts them into the instantiation context $[[Θ]]$
(specifying that they must be instantiated in the extended context
$[[Γ, pbs]]$ or $[[Γ, nbs]]$) and substitute the quantified variables
for them in the input type.

@@

\paragraph{Functions}
To infer the subtyping of the function types, the algorithm
makes two calls: 
\begin{enumerate*}
  \item[(i)] a recursive call ensuring the subtyping of the result types, and
  \item[(ii)] a call to positive subtyping (or rather super-typing) on the argument types.
\end{enumerate*}
The resulting constraints are merged
using a special procedure defined in \cref{sec:constraint-merge}
and returned as the output.

\paragraph{Algorithmic variables}
If one of the sides of the subtyping is a unification variable, the algorithm
creates a new constraint. Because the right-hand side of the subtyping is always
declarative, it is only the left-hand side that can be a unification variable.
Moreover, another invariant we preserve prevents the negative algorithmic
variables from occurring in types during the negative subtyping algorithm. It
means that the only possible form of the subtyping here is $[[α̂⁺]] [[≥]]
[[iP]]$, which is covered by \ruleref{\ottdruleAPUVarLabel}.

@@

\paragraph{Variables}
  The variable rules \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel}
  are trivial: as the input types do not have algorithmic variables, and are already equal, 
  the unification returns no constraints.

\paragraph{Shifts}
  The shift rules \ruleref{\ottdruleUShiftDLabel} and \ruleref{\ottdruleUShiftULabel}
  require the two input types to be formed by the same shift constructor. 
  They remove this constructor, unify the types recursively, and return the resulting
  set of constraints.

\paragraph{Quantifiers} 
  Similarly, the quantifier rules \ruleref{\ottdruleUForallLabel} and \ruleref{\ottdruleUExistsLabel}
  require the quantifier variables on the left-hand side and the right-hand side to be the same.
  This requirement is complete because we assume the input types of the unification 
  to be normalized, and thus, the equivalence implies alpha-equivalence. 
  In the implementation of this rule, an alpha-renaming might be needed to ensure 
  that the quantified variables are the same, however, we omit it for brevity.

\paragraph{Functions}
  Rule \ruleref{\ottdruleUArrowLabel} unifies two functional types. 
  First, it unifies their argument types and their result types recursively.
  Then it merges the resulting constraints using the procedure described in \cref{sec:constraint-merge}.

@@

\paragraph{Algorithmic variable}
  Finally, if the left-hand side of the unification is an algorithmic variable,
  \ruleref{\ottdruleUNVarLabel} or \ruleref{\ottdruleUPVarLabel} is applied. 
  It simply checks that the right-hand side type is well-formed in the required
  instantiation context, and returns a newly created constraint restricting the variable
  to be equal to the right-hand side type.

@@

\paragraph{Merging matching constraint entries}
Two \emph{matching} entries formed in the same context $[[Γ]]$ 
can be merged as shown in \cref{fig:merge-entries}.
Suppose that $[[scE1]]$ and $[[scE2]]$ are input entries. 
The result of the merge $[[scE1 & scE2]]$ is 
\emph{the weakest entry which implies both $[[scE1]]$ and $[[scE2]]$}.

@@

\paragraph{Merging constraint sets}
  The algorithm for merging constraint sets is shown in \cref{fig:merge-subtyping-constraints}.
  As discussed, the result of merge $[[SC1]]$ and $[[SC2]]$ consists of three parts: 
  \begin{enumerate*}
    \item[(i)] the entries of $[[SC1]]$ that do not match any entry of $[[SC2]]$;
    \item[(ii)] the entries of $[[SC2]]$ that do not match any entry of $[[SC1]]$; and
    \item[(iii)] the merge (\cref{fig:merge-entries}) of matching entries.
  \end{enumerate*}

@@

\paragraph{The Type Upgarde}
The type upgrade algorithm uses the least upper bound algorithm as a subroutine.
It exploits the idea that the free variables of a positive type $[[iQ]]$
cannot disappear in its subtypes (see \cref{prop:subtyping-preserves-fv}). 
It means that if 
a type $[[iP]]$ has free variables not occurring 
in $[[iP']]$, then any common supertype of $[[iP]]$
and $[[iP']]$ must not contain these variables either.
This way, any supertype of $[[iP]]$
not containing certain variables $[[pnas]]$ must also be 
a supertype of $[[iP' = [pnbs/pnas]iP ]]$, where $[[pnbs]]$ are fresh;
and vice versa: any common supertype of $[[iP]]$ and $[[iP']]$
does not contain $[[pnas]]$ nor $[[pnbs]]$.

@@

\paragraph{The Least Upper Bound}
The Least Upper Bound algorithm we use operates on \emph{positive}
types. This way, the inference rules of the algorithm
analyze the three possible shapes of the input types:
a variable type, an existential type, and a shifted computation.

@@

\paragraph{Variables} 
  Rules \ruleref{\ottdruleAUPVarLabel} and
  \ruleref{\ottdruleAUNVarLabel} generalize two equal variables.
  In this case, the resulting pattern is the variable itself,
  and no `holes' are needed.

\paragraph{Shifts} 
  Rules \ruleref{\ottdruleAUShiftDLabel} and
  \ruleref{\ottdruleAUShiftULabel} operate by congruence:
  they anti-unify the bodies of the shifts recursively and add
  the shift constructor back to the resulting pattern.

\paragraph{Quantifiers}
  Rules \ruleref{\ottdruleAUForallLabel} and
  \ruleref{\ottdruleAUExistsLabel} are symmetric. 
  They generalize two quantified types congruently, 
  similarly to the shift rules. 
  However, we also require that the quantified variables
  are fresh, and that the left-hand side variables are 
  equal to the corresponding variables on the right-hand side.
  To ensure it, alpha-renaming might be required in the
  implementation.

@@

\paragraph{Functions}
  Rule \ruleref{\ottdruleAUArrowLabel} congruently generalizes two function types. 
  An arrow type is the only binary constructor, 
  and thus, it is the only rule where the union of the anti-unification results is substantial.
  The interesting is the case when the resulting generalization of the 
  input types and the resulting generalization of the output types 
  have `holes' mapped to the same pair of types. 
  In this case, the algorithm must merge the `holes' into one.
  For example, the anti-unification of 
  $[[↓α⁻ → α⁻]]$ and $[[↓β⁻ → β⁻]]$
  must result in $[[↓γ̂⁻ → γ̂⁻]]$,
  rather than $[[↓γ1̂⁻ → γ2̂⁻]]$.

@@

\paragraph{The Anti-Unification Rule}
  Rule \ruleref{\ottdruleAUAULabel} is the base case of the anti-unification
  algorithm. If the congruent rules are not applicable, 
  it means that the input types have a substantially different structure,
  and thus, the only option is to create a `hole'. 
  There are three important aspects of this rule that we would like to discuss.

@@

\paragraph{Variables}
  Rule \ruleref{\ottdruleATVarLabel} 
  infers the type of a positive variable by looking it up in the 
  term variable context and normalizing the result.

\paragraph{Annotations}
  Rules \ruleref{\ottdruleATPAnnotLabel} and \ruleref{\ottdruleATNAnnotLabel}
  are symmetric.
  First, they check that the annotated type is well-formed in the 
  given context $[[Γ]]$. Then they make a recursive call to infer the 
  type of annotated expression, check that the inferred type is a subtype of 
  the annotation, and return the normalized annotation.

\paragraph{Abstractions}
  Rule \ruleref{\ottdruleATtLamLabel} infers the type of a lambda abstraction.
  It checks the well-formedness of the annotation $[[iP]]$,
  makes a recursive call to infer the type of the body in the extended context, 
  and returns the corresponding arrow type.
  Since the annotation $[[iP]]$ is allowed to be non-normalized,
  the rule also normalizes the resulting type.

@@

\paragraph{Return and Thunk}
  Rules \ruleref{\ottdruleATThunkLabel} and \ruleref{\ottdruleATReturnLabel}
  are similar to the declarative rules: they make a recursive call
  to type the body of the thunk or the return expression and
  put the shift on top of the result.

\paragraph{Unpack}
  Rule \ruleref{\ottdruleATUnpackLabel}
  allows one to unpack an existential type.
  First, it infers the existential type $[[∃nas.iP]]$ of the value being unpacked,
  and since the type is guaranteed to be normalized, binds 
  the quantified variables with $[[nas]]$.
  Then it infers the type of the body in the appropriately extended context
  and checks that the inferred type does not depend on $[[nas]]$
  by checking well-formedness $[[Γ ⊢ iN]]$.

\paragraph{Let Binders}
  Rule \ruleref{\ottdruleATVarLetLabel} represents the type inference of a
  standard let binder. It infers the type of the bound value $[[v]]$, and makes
  a recursive call to infer the type of the body in the extended context.

@@

\paragraph{Application to an Empty List of Arguments}
  Rule \ruleref{\ottdruleATEmptyAppLabel}
  is the base case of application inference. 
  If the list of applied arguments is empty, 
  the inferred type is the type of the head,
  and the algorithm returns it after normalizing.

\paragraph{Application of a Polymorphic Type $[[∀]]$}
  Rule \ruleref{\ottdruleATForallAppLabel},
  analogously to the declarative case,
  is the rule ensuring the implicit elimination of the universal quantifiers. 
  This is the place where the algorithmic variables are introduced.
  The algorithm simply replaces the quantified variables 
  $[[pas]]$ with fresh algorithmic variables $[[puas]]$,
  and makes a recursive call in the extended context.

@@

\paragraph{Application of an Arrow Type}
  Rule \ruleref{\ottdruleATArrowAppLabel}
  is the main rule of algorithmic application inference.
  It is applied when the head has an arrow type $[[uQ → uN]]$.
  First, it infers the type of the first argument $[[v]]$,
  and then, calling the algorithmic subtyping,
  finds $[[SC1]]$---the \emph{minimal} constraint ensuring that 
  $[[uQ]]$ is a supertype of the type of $[[v]]$.
  Then it makes a recursive call applying $[[uN]]$ to the rest of the arguments 
  and merges the resulting constraint with $[[SC1]]$.

>>>>

\paragraph*{Algorithmic Variables}
To design a mode-correct inference system, we slightly modify the language we operate on.
The entities (terms, types, contexts) that the algorithm manipulates we call \emph{algorithmic}. 
They extend the previously defined declarative terms and types by adding 
\emph{algorithmic type variables} (\aka unification variables). 
The algorithmic variables represent unknown types, 
which cannot be inferred immediately but are promised to be instantiated
as the algorithm proceeds.

@@

\paragraph*{Algorithmic Types}
The syntax of algorithmic types extends the declarative syntax by adding
algorithmic variables as new terminals. We add positive algorithmic variables $[[α̂⁺]]$ 
to the syntax of positive types, and negative algorithmic variables $[[α̂⁻]]$ to the 
syntax of negative types. All the constructors of the system can be applied 
to \emph{algorithmic} types, however, algorithmic variables cannot be abstracted by the
quantifiers $[[∀]]$ and $[[∃]]$.

\paragraph*{Algorithmic Contexts $[[Ξ]]$ and Well-formedness}
To specify when algorithmic types are well-formed, we define algorithmic
contexts $[[Ξ]]$ as sets of algorithmic variables. Then
$[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ represent the well-formedness judgment of
algorithmic terms defined as expected. Informally, they check that all free
declarative variables are in $[[Γ]]$, and all free algorithmic variables are in
$[[Ξ]]$. Most of the rules are inherited from the well-formedness of
\emph{declarative} types: the declarative variables are checked to belong to the
context $[[Γ]]$, the quantifiers extend the context, type constructors are
well-formed congruently. As we extend the syntax with algorithmic variables, we
also add two base-case rules for them: \ruleref{\ottdruleWFATPUVarLabel} and
\ruleref{\ottdruleWFATNUVarLabel} (see \cref{fig:algo-wf}).

@@

\paragraph*{Instantiation Context $[[Θ]]$}
When one instantiates an algorithmic variable, one may only use type variables
\emph{available in its scope}. As such, each algorithmic variable must remember
the context at the moment when it was introduced. In our algorithm, this
information is represented by an \emph{instantiation context} $[[Θ]]$---a set of
pairs associating algorithmic variables and declarative contexts.

\paragraph*{Algorithmic Substitution}
We define the algorithmic substitution $[[uσ]]$ as a mapping from algorithmic
variables to \emph{declarative} types. The signature $[[Θ ⊢ uσ : Ξ]]$
specifies the domain and the range of $[[uσ]]$: for each variable $[[α̂±]]$
in $[[Ξ]]$, there exists an corresponding entry in $[[Θ]]$ associating 
$[[α̂±]]$ with a declarative context $[[Γ]]$ such that $[[ [uσ]α̂± ]]$
is well-formed in $[[Γ]]$. In addition, we assume that $[[uσ]]$
acts as the identity on the variables not in $[[Ξ]]$.

\paragraph*{Algorithmic Normalization}
Similarly to well-formedness, the normalization of algorithmic types is defined
by extending the declarative definition (\cref{fig:type-nf}) with the
algorithmic variables. To the rules repeating the declarative normalization, we
add rules saying that normalization is trivial on algorithmic variables.

@@

\paragraph*{Auxiliary Functions}
\begin{itemize}
  \item
    We define $[[dom(SC)]]$---the domain of constraint set $[[SC]]$---as a set of
    algorithmic variables that it restricts. Similarly, we define $[[dom(Θ)]]$---the
    domain of instantiation context---as a set of algorithmic variables that
    $[[Θ]]$ associates with their contexts.  
  \item We write $[[Θ(α̂±)]]$ to denote the
    declarative context associated with $[[α̂±]]$ in $[[Θ]]$. 
  \item $[[uv(uN)]]$ and $[[uv(uP)]]$ denote the set of free algorithmic variables in $[[uN]]$ and
    $[[uP]]$ respectively.
  \item We write $[[Γ ⊢ Θ]]$ to denote that each
    declarative context associated with an algorithmic variable in $[[Θ]]$ is 
    a subcontext (subset) of $[[Γ]]$.
\end{itemize}

\paragraph*{Equivalent Substitutions}
In the proofs of the algorithm correctness, 
we often state or require that two substitutions are equivalent
on a given set of algorithmic variables.
We denote it as $[[Θ ⊢ uσ' ≈ uσ : Ξ]]$ meaning that
for each $[[α̂±]]$ in $[[Ξ]]$, substitutions $[[uσ]]$ and $[[uσ']]$
map $[[α̂±]]$ to types equivalent in the corresponding context:
$[[ Θ(α̂±) ⊢ [uσ']α̂± ≈ [uσ]α̂± ]]$.

\paragraph*{Constraint well-formedness and satisfaction}
Suppose that $[[Θ]]$ is an instantiation context. 
We say that constraint set $[[SC]]$ is well-formed in $[[Θ]]$ 
(denoted as $[[Θ ⊢ SC]]$) 
if for every every constrain entry $[[scE ∊ SC]]$
associating a variable $[[α̂±]]$ with a type $[[iP]]$,
this type is well-formed in the corresponding declarative context $[[Θ(α̂±)]]$.

@@

\paragraph*{Variables} Rules \ruleref{\ottdruleANVarLabel} and \ruleref{\ottdruleAPVarLabel}
say that if both of the input types are equal declarative variables,
they are subtypes of each other, with no constraints (as there are no algorithmic variables).

\paragraph*{Shifts} Rules \ruleref{\ottdruleAShiftDLabel} and
\ruleref{\ottdruleAShiftULabel} cover the downshift and the upshift cases,
respectively. If the input types are constructed by shifts, then the subtyping
can only hold if they are equivalent. This way, the algorithm must find the
instantiations of the algorithmic variables on the left-hand side
such that these instantiations make the left-hand side and the ride-hand side
equivalent. For this purpose, the algorithm invokes the
unification procedure (\cref{sec:unification}) preceded by the normalization of the input types.
It returns the resulting constraints given by the unification algorithm.

\paragraph*{Quantifiers}  
Rules \ruleref{\ottdruleAForallLabel} and 
\ruleref{\ottdruleAExistsLabel} are symmetric. 
Declaratively, the quantified variables on the left-hand side must 
be instantiated with types, which are not known beforehand.
We address this problem by algorithmization 
of the quantified variables (see \cref{sec:algo-syntax}).
The rule introduces fresh algorithmic variables
$[[puas]]$ or $[[nuas]]$,
puts them into the instantiation context $[[Θ]]$
(specifying that they must be instantiated in the extended context
$[[Γ, pbs]]$ or $[[Γ, nbs]]$) and substitute the quantified variables
for them in the input type.

@@

\paragraph*{Functions}
To infer the subtyping of the function types, the algorithm
makes two calls: 
\begin{enumerate*}
  \item[(i)] a recursive call ensuring the subtyping of the result types, and
  \item[(ii)] a call to positive subtyping (or rather super-typing) on the argument types.
\end{enumerate*}
The resulting constraints are merged
using a special procedure defined in \cref{sec:constraint-merge}
and returned as the output.

\paragraph*{Algorithmic variables}
If one of the sides of the subtyping is a unification variable, the algorithm
creates a new constraint. Because the right-hand side of the subtyping is always
declarative, it is only the left-hand side that can be a unification variable.
Moreover, another invariant we preserve prevents the negative algorithmic
variables from occurring in types during the negative subtyping algorithm. It
means that the only possible form of the subtyping here is $[[α̂⁺]] [[≥]]
[[iP]]$, which is covered by \ruleref{\ottdruleAPUVarLabel}.

@@

\paragraph*{Variables}
  The variable rules \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel}
  are trivial: as the input types do not have algorithmic variables, and are already equal, 
  the unification returns no constraints.

\paragraph*{Shifts}
  The shift rules \ruleref{\ottdruleUShiftDLabel} and \ruleref{\ottdruleUShiftULabel}
  require the two input types to be formed by the same shift constructor. 
  They remove this constructor, unify the types recursively, and return the resulting
  set of constraints.

\paragraph*{Quantifiers} 
  Similarly, the quantifier rules \ruleref{\ottdruleUForallLabel} and \ruleref{\ottdruleUExistsLabel}
  require the quantifier variables on the left-hand side and the right-hand side to be the same.
  This requirement is complete because we assume the input types of the unification 
  to be normalized, and thus, the equivalence implies alpha-equivalence. 
  In the implementation of this rule, an alpha-renaming might be needed to ensure 
  that the quantified variables are the same, however, we omit it for brevity.

\paragraph*{Functions}
  Rule \ruleref{\ottdruleUArrowLabel} unifies two functional types. 
  First, it unifies their argument types and their result types recursively.
  Then it merges the resulting constraints using the procedure described in \cref{sec:constraint-merge}.

@@

\paragraph*{Algorithmic variable}
  Finally, if the left-hand side of the unification is an algorithmic variable,
  \ruleref{\ottdruleUNVarLabel} or \ruleref{\ottdruleUPVarLabel} is applied. 
  It simply checks that the right-hand side type is well-formed in the required
  instantiation context, and returns a newly created constraint restricting the variable
  to be equal to the right-hand side type.

@@

\paragraph*{Merging matching constraint entries}
Two \emph{matching} entries formed in the same context $[[Γ]]$ 
can be merged as shown in \cref{fig:merge-entries}.
Suppose that $[[scE1]]$ and $[[scE2]]$ are input entries. 
The result of the merge $[[scE1 & scE2]]$ is 
\emph{the weakest entry which implies both $[[scE1]]$ and $[[scE2]]$}.

@@

\paragraph*{Merging constraint sets}
  The algorithm for merging constraint sets is shown in \cref{fig:merge-subtyping-constraints}.
  As discussed, the result of merge $[[SC1]]$ and $[[SC2]]$ consists of three parts: 
  \begin{enumerate*}
    \item[(i)] the entries of $[[SC1]]$ that do not match any entry of $[[SC2]]$;
    \item[(ii)] the entries of $[[SC2]]$ that do not match any entry of $[[SC1]]$; and
    \item[(iii)] the merge (\cref{fig:merge-entries}) of matching entries.
  \end{enumerate*}

@@

\paragraph*{The Type Upgarde}
The type upgrade algorithm uses the least upper bound algorithm as a subroutine.
It exploits the idea that the free variables of a positive type $[[iQ]]$
cannot disappear in its subtypes (see \cref{prop:subtyping-preserves-fv}). 
It means that if 
a type $[[iP]]$ has free variables not occurring 
in $[[iP']]$, then any common supertype of $[[iP]]$
and $[[iP']]$ must not contain these variables either.
This way, any supertype of $[[iP]]$
not containing certain variables $[[pnas]]$ must also be 
a supertype of $[[iP' = [pnbs/pnas]iP ]]$, where $[[pnbs]]$ are fresh;
and vice versa: any common supertype of $[[iP]]$ and $[[iP']]$
does not contain $[[pnas]]$ nor $[[pnbs]]$.

@@

\paragraph*{The Least Upper Bound}
The Least Upper Bound algorithm we use operates on \emph{positive}
types. This way, the inference rules of the algorithm
analyze the three possible shapes of the input types:
a variable type, an existential type, and a shifted computation.

@@

\paragraph*{Variables} 
  Rules \ruleref{\ottdruleAUPVarLabel} and
  \ruleref{\ottdruleAUNVarLabel} generalize two equal variables.
  In this case, the resulting pattern is the variable itself,
  and no `holes' are needed.

\paragraph*{Shifts} 
  Rules \ruleref{\ottdruleAUShiftDLabel} and
  \ruleref{\ottdruleAUShiftULabel} operate by congruence:
  they anti-unify the bodies of the shifts recursively and add
  the shift constructor back to the resulting pattern.

\paragraph*{Quantifiers}
  Rules \ruleref{\ottdruleAUForallLabel} and
  \ruleref{\ottdruleAUExistsLabel} are symmetric. 
  They generalize two quantified types congruently, 
  similarly to the shift rules. 
  However, we also require that the quantified variables
  are fresh, and that the left-hand side variables are 
  equal to the corresponding variables on the right-hand side.
  To ensure it, alpha-renaming might be required in the
  implementation.

@@

\paragraph*{Functions}
  Rule \ruleref{\ottdruleAUArrowLabel} congruently generalizes two function types. 
  An arrow type is the only binary constructor, 
  and thus, it is the only rule where the union of the anti-unification results is substantial.
  The interesting is the case when the resulting generalization of the 
  input types and the resulting generalization of the output types 
  have `holes' mapped to the same pair of types. 
  In this case, the algorithm must merge the `holes' into one.
  For example, the anti-unification of 
  $[[↓α⁻ → α⁻]]$ and $[[↓β⁻ → β⁻]]$
  must result in $[[↓γ̂⁻ → γ̂⁻]]$,
  rather than $[[↓γ1̂⁻ → γ2̂⁻]]$.

@@

\paragraph*{The Anti-Unification Rule}
  Rule \ruleref{\ottdruleAUAULabel} is the base case of the anti-unification
  algorithm. If the congruent rules are not applicable, 
  it means that the input types have a substantially different structure,
  and thus, the only option is to create a `hole'. 
  There are three important aspects of this rule that we would like to discuss.

@@

\paragraph*{Variables}
  Rule \ruleref{\ottdruleATVarLabel} 
  infers the type of a positive variable by looking it up in the 
  term variable context and normalizing the result.

\paragraph*{Annotations}
  Rules \ruleref{\ottdruleATPAnnotLabel} and \ruleref{\ottdruleATNAnnotLabel}
  are symmetric.
  First, they check that the annotated type is well-formed in the 
  given context $[[Γ]]$. Then they make a recursive call to infer the 
  type of annotated expression, check that the inferred type is a subtype of 
  the annotation, and return the normalized annotation.

\paragraph*{Abstractions}
  Rule \ruleref{\ottdruleATtLamLabel} infers the type of a lambda abstraction.
  It checks the well-formedness of the annotation $[[iP]]$,
  makes a recursive call to infer the type of the body in the extended context, 
  and returns the corresponding arrow type.
  Since the annotation $[[iP]]$ is allowed to be non-normalized,
  the rule also normalizes the resulting type.

@@

\paragraph*{Return and Thunk}
  Rules \ruleref{\ottdruleATThunkLabel} and \ruleref{\ottdruleATReturnLabel}
  are similar to the declarative rules: they make a recursive call
  to type the body of the thunk or the return expression and
  put the shift on top of the result.

\paragraph*{Unpack}
  Rule \ruleref{\ottdruleATUnpackLabel}
  allows one to unpack an existential type.
  First, it infers the existential type $[[∃nas.iP]]$ of the value being unpacked,
  and since the type is guaranteed to be normalized, binds 
  the quantified variables with $[[nas]]$.
  Then it infers the type of the body in the appropriately extended context
  and checks that the inferred type does not depend on $[[nas]]$
  by checking well-formedness $[[Γ ⊢ iN]]$.

\paragraph*{Let Binders}
  Rule \ruleref{\ottdruleATVarLetLabel} represents the type inference of a
  standard let binder. It infers the type of the bound value $[[v]]$, and makes
  a recursive call to infer the type of the body in the extended context.

@@

\paragraph*{Application to an Empty List of Arguments}
  Rule \ruleref{\ottdruleATEmptyAppLabel}
  is the base case of application inference. 
  If the list of applied arguments is empty, 
  the inferred type is the type of the head,
  and the algorithm returns it after normalizing.

\paragraph*{Application of a Polymorphic Type $[[∀]]$}
  Rule \ruleref{\ottdruleATForallAppLabel},
  analogously to the declarative case,
  is the rule ensuring the implicit elimination of the universal quantifiers. 
  This is the place where the algorithmic variables are introduced.
  The algorithm simply replaces the quantified variables 
  $[[pas]]$ with fresh algorithmic variables $[[puas]]$,
  and makes a recursive call in the extended context.

@@

\paragraph*{Application of an Arrow Type}
  Rule \ruleref{\ottdruleATArrowAppLabel}
  is the main rule of algorithmic application inference.
  It is applied when the head has an arrow type $[[uQ → uN]]$.
  First, it infers the type of the first argument $[[v]]$,
  and then, calling the algorithmic subtyping,
  finds $[[SC1]]$---the \emph{minimal} constraint ensuring that 
  $[[uQ]]$ is a supertype of the type of $[[v]]$.
  Then it makes a recursive call applying $[[uN]]$ to the rest of the arguments 
  and merges the resulting constraint with $[[SC1]]$.

### 2024-06-11 paper/tex/05-proof.tex
amend
corrections

\paragraph{Determinacy}
    One of the properties that our proof relies on is the determinacy of the 
    typing algorithm: the output (the inferred type)
    is uniquely determined by the input (the term and the contexts).
    Determinacy is not hard to demonstrate by structural induction: 
    in every algorithmic inference system, only one inference rule can be applied 
    for a given input.  However, we need to prove it for \emph{every} 
    subroutine of the algorithm. Ultimately, it requires 
    the determinacy of such procedures as \emph{generation of fresh variables}, 
    which is easy to ensure, but must be taken into account in the implementation.

@@

\paragraph{Mutuality of the Soundness and Completeness Proofs}

@@

The only way to do so is by applying the soundness of minimal instantiation
(\cref{lemma:minimal-instantiation-soundness}). However, first, the declarative
tree must be converted to the corresponding algorithmic one (by completeness!).
This way, the soundness and completeness proofs are mutually dependent.

\paragraph{Inductive Metric}
    The soundness and completeness lemmas are proven by \emph{mutual} induction. 
    Since the declarative and the algorithmic systems do not depend on each other,
    we must introduce a uniform \emph{metric} on which the induction is conducted.
    We define the metric gradually, starting from the
    auxiliary function---the size of a judgment $\size{J}$.

>>>>

\paragraph*{Determinacy}
    One of the properties that our proof relies on is the determinacy of the 
    typing algorithm: the output (the inferred type)
    is uniquely determined by the input (the term and the contexts).
    Determinacy is not hard to demonstrate by structural induction: 
    in every algorithmic inference system, only one inference rule can be applied 
    for a given input.  However, we need to prove it for \emph{every} 
    subroutine of the algorithm. Ultimately, it requires 
    the determinacy of such procedures as \emph{generation of fresh variables}, 
    which is easy to ensure, but must be taken into account in the implementation.

@@

\paragraph*{Mutuality of the Soundness and Completeness Proofs}

@@

The only way to do so is by applying the soundness of minimal instantiation
(see the lemma above). However, first, the declarative
tree must be converted to the corresponding algorithmic one (by completeness!).
This way, the soundness and completeness proofs are mutually dependent.

\paragraph*{Inductive Metric}
    The soundness and completeness lemmas are proven by \emph{mutual} induction. 
    Since the declarative and the algorithmic systems do not depend on each other,
    we must introduce a uniform \emph{metric} on which the induction is conducted.
    We define the metric gradually, starting from the
    auxiliary function---the size of a judgment $\size{J}$.
